### 2014-09-05 relax2.tex
relaxII

We start with a branch event $e_b$ in the trace, without loss of generality, suppose its true branch is taken, we need to explore the false branch (thread-locally with the same thread as $e_b$) and add constraints to model it. As the static analysis has some known challenges, we make several simplifying assumptions: (1) the unexplored branch should not contain loops or recursions because the loop bound may depend on some shared variable which is statically unknown; (2) every base reference variable should be fully resolved to a concrete object, as required by our analysis (Section~\ref{sec:relax1}). We apply dataflow analysis (reaching definitions) to judge: If the base reference depends on some shared read in the explored branch, of which the value is statically unknown, it cannot be statically resolved to an object; If the base reference depends on multiple definitions from different paths, it cannot be statically resolved to an object. Otherwise, it can be resolved. (3) We assume the unexplored branch does not contain array access because the array accesses often use the index expressions that are statically unknown. Similarly, as the collections often use array for implementation, we assume they are also absent. (4) We inline the method calls (for two levels) in the unexplored branch, similar to the base object, we assume the caller object can be fully resolved statically.  If any of the above assumptions is violated, we  avoid exploring the unexplored branches and adopt the purely dynamic solution in Section~\ref[sec:relax1}. 
We use the example in Figure~\ref{fig:running} for demonstration.

@@

\begin{figure}
\centering
\begin{tabular}{l}
 {\tt 4: if($x^5$<10)}  \\ %{\tt if($x^5$<10)}\\
{\tt 7: $W^7_{S}$=0;}   \\%{\tt $W^7_{S}$=0;}\\
 {{\tt 8:  else}}   \\ %{\color{Gray}{\tt if($x^5\geq$10)}}\\
 {{\tt 9: \ \ $z^9$=$R^{9}_{o1.f}$;}}  \\ %{\color{Gray}{\tt $z^{9}$=$R^{9}_{o1.f}$;}}\\
 {{\tt 10: \ if($x^5$<$z^9$)}}  \\%{\color{Gray}{\tt if($x^5<z^{9}$)}}\\
 {{\tt 11: \ \ $W^{11}_{S}$=0;}} \\%{\color{Gray}{\tt $W^{11}_{S}$=0;}}\\ 
 {{\tt 12: \ else }}  \\%{\color{Gray}{\tt if($x^5\geq z^{10}$)}}\\
 {{\tt 13: \ \ $W^{13}_{S}$=$l^{id}_{o2.f}$; }}   \\%{\color{Gray}{\tt $W^{13}_{S}$=$l^{id}_{o2.f}$;}}\\
\end{tabular}
\caption{Running Example (shared variables are in bold font). }
\label{fig:pathssa}
\end{figure}

@@

{\bf Race Condition Constraint\ }  The first constraint is the race condition constraint, e.g., $O'_{4}=O'(e_11)$ specifies that the event at line 4 may form a race with the event at line 11, which is not executed in the original trace.

{\bf Intra-thread Control Flow Constraint \ } Different from the strategy in Section~\ref{sec:relax1}, we do not assert the branch the predicted run will take. Instead, we assert that all branches inside the branch structure may be taken, which is trivially $true$. The solver automatically reasons about which branch should be taken, as indicated in the solution. Note that only one of the two branches can be taken because the conditions for them contradict.

{\bf Intra-thread Value Constraint \ } Intra-thread value constraint is the core constraint that realizes the exploration different branches. It asserts the value constraints imposed by the {\sf assign/heapr/heapw} events take effect only when the guarding condition is satisfied. More formally, we have

\[ cons(n) = \left\{ 
  \begin{array}{l l}
  (\bigwedge_{n_1\in n_{if}}{  (expr(header_n) \rightarrow  cons(n_1))}) \\
  \bigwedge (\bigwedge_{n_2\in n_{else}}  { (\neg expr(header_n) \rightarrow  cons(n_2))}) \\
   \quad  \quad  \quad \quad  \quad  \quad  \quad  \quad  \quad  \text{if n is a branch} \\
  equation(n)      \\                                  
   \quad  \quad  \quad \quad  \quad  \quad \quad  \quad  \quad  \text{otherwise} \\ 
  \end{array} \right.\]
Here, $expr(header)$ is the helper function that returns the condition associated with the header. $equation(n)$ returns the equation inside each event, which imposes the value constraint.  In the first case when the instruction $n$ is a branch, we assert the value constraint imposed by each instruction in the if-branch can take effect only when the if-branch is taken, i.e., when $expr(header_n)$ is evaluated as true,  and the value constraint imposed by instruction in the else-branch can take effect only when the else-branch is taken. In the second case, when the instruction $n$ is a normal instruction that contains the equation, we assert the equation as the value constraint. Note that the constraint is defined recursively because a branch  may contain another branch structure.

Consider the example in Figure~\ref{fig:pathssa},  the constraint is,
$$
\begin{array}{l}
&(\theta'(x^5)<10 \rightarrow \theta'(W^7_{S})=0) \wedge (\theta'(x^5)\geq 10 \rightarrow \theta'(z^9)=\theta'(R^{9}_{o1.f}) 
&\wedge  (\theta'(x^5)\geq 10 \rightarrow  
& \ \ \ \ \  \ \ (\theta'(x^5)<\theta'(z^9) \rightarrow \theta'(W^{11}_{S})=0 
& \ \ \ \ \  \ \ \wedge \theta'(x^5)\geq \theta'(z^9) \rightarrow \theta'(W^{13}_{S})=\theta'(l^{id}_{o2.f}))
&)\\ 
\end{array}
$$

{\bf Phi node}

{\bf Intra-thread Order Constraints\ } The intra-thread order constraint captures the program order imposed by the branch structure and the instruction sequence. More formally, for each branch $n$, we have,

\[ cons(n) = \left\{ 
  \begin{array}{l l}
  (\bigwedge_{n_1\in n_{if}}{O'(header_n)<O'(n_1)<O'(end_n)}) \\
   (\bigwedge_{n_1 and n_3 are adjacent in n_{if}}{O'(n_1)<O'(n_3)}) \\
   (\bigwedge_{n_2\in n_{else}}{O'(header_n)<O'(n_2)<O'(end_n)}) \\
    (\bigwedge_{n_2 and n_4 are adjacent in n_{else}}{O'(n_2)<O'(n_4)}) \\
  \end{array} \right.\]
Here, $end_n$ is a node that denotes the first event after the branch.  The order constraints are imposed by the program, and therefore, are unconditionally valid. That is why we do not specify the order constraints under specific branch conditions.

{\bf Inter-thread value constraints\ } The inter-thread value constraints are identical to those in Section~\ref{sec:relax1}. Interesting, we do not need to specify the value constraints under the specific branch condition. The underlying reason is that, the intra-thread value constraint already encodes the branch condition information, if the branch condition is not satisfied at runtime, the value constraint does not take effect and the read of a shared variable cannot propagate its value downstream, even if it is assigned to a value.

>>>>

We start with a branch event $e_b$ in the trace, without loss of generality, suppose its true branch is taken, we need to explore the false branch (thread-locally with the same thread as $e_b$) and add constraints to model it. As the static analysis has some known challenges, we make several simplifying assumptions: (1) the unexplored branch should not contain loops or recursions because the loop bound may depend on some shared variable which is statically unknown; (2) every base reference variable should be fully resolved to a concrete object, as required by our analysis (Section~\ref{sec:relax1}). We apply dataflow analysis (reaching definitions) to judge: If the base reference depends on some shared read in the explored branch, of which the value is statically unknown, it cannot be statically resolved to an object; If the base reference depends on multiple definitions from different paths, it cannot be statically resolved to an object. Otherwise, it can be resolved. (3) We assume the unexplored branch does not contain array access because the array accesses often use the index expressions that are statically unknown. Similarly, as the collections often use array for implementation, we assume they are also absent. (4) We inline the method calls (for two levels) in the unexplored branch, similar to the base object, we assume the caller object can be fully resolved statically.  If any of the above assumptions is violated, we  avoid exploring the unexplored branches and adopt the purely dynamic solution in Section~\ref{sec:relax1}. 
We use the example in Figure~\ref{fig:running} for demonstration.

@@

\begin{figure}
\centering
\begin{tabular}{l}
 {\tt 4: if($R^4_y$>2)}  \\ %{\tt if($x^5$<10)}\\
{\tt 5: \ \ $z^5$=1+$R^5_x$;}   \\%{\tt $W^7_{S}$=0;}\\
 {{\tt 6:  else}}   \\ %{\color{Gray}{\tt if($x^5\geq$10)}}\\
 {{\tt 7: \ \ $w^7$=$2+R^{7}_{x}$;}}  \\ %{\color{Gray}{\tt $z^{9}$=$R^{9}_{o1.f}$;}}\\
\end{tabular}
\caption{SSA form of both true and false branches. }
\label{fig:pathssa}
\end{figure}

@@

{\bf Race Condition Constraint\ }  The first type of constraint is the race condition constraint, e.g., $O'_{e_2}=O'(e_7)$ specifies that the event at line 2 may form a race with the event at line 7, which is actually not executed in the original trace.

{\bf Intra-thread Control Flow Constraint \ } Different from the strategy in Section~\ref{sec:relax1}, we do not assert the branch the predicted run will take. Instead, we assert that both branches are possible, which is always $true$. The solver automatically reasons about which branch should be taken and encodes it in the solution. Note that only one of the two branches can be taken because the conditions for the branches contradict.

{\bf Intra-thread Value Constraint \ } Intra-thread value constraint is the core constraint that realizes the exploration different branches. It asserts the value constraints imposed by the {\sf assign/heapr/heapw} events take effect only when the guarding condition is satisfied. 
Suppose $IF(e_b)$ denotes the event sequence that corresponds to the true branch, $ELSE(e_b)$ denotes the event sequence that corresponds to the false branch, the branch event $e_b$ is  $if(x_b<y_b)$, the event $e_m$ in $IF(e_b)$ is in the form of $x_m=y_m+z_m$ and the event $e_n$ in $ELSE(e_b)$ is in the form of $x_n=y_n+z_n$.

Consider our example, the corresponding value constraint is, 
$$
\begin{array}{l}
\theta'(R^4_y)>2 \rightarrow \theta'(z^5)=1+\theta'(R^5_x) \\ 
\wedge  \\
\theta'(R^4_y)\leq 2 \rightarrow  \theta'(w^7)=2+\theta'(R^{7}_{x})\\
\end{array}
$$

{\bf Assignment with {\sf Phi}\ }  For the assignment with the Phi, e.g., $y=Phi(y_1, x>0, y_2, x\leq 0)$, the encoding is similar.
$$
\begin{array}{l}
\theta'(x)>0 \rightarrow \theta'(y)=\theta'(y_1) \wedge \theta'(x)\leq 0 \rightarrow \theta'(y)=\theta'(y_2)
\end{array}
$$

In case that the unexplored branch contains another branch structure, the above encoding process is applied recursively and the resultant constraints are nested.

{\bf Intra-thread Order Constraints\ } The intra-thread order constraint captures the program order imposed by the branch structure and the instruction sequence. Suppose $IF(e_b)$=$e_{i_1}e_{i_2}\dots e_{i_m}$, $ELSE(e_b)=e_{j_1}e_{j_2}\dots e_{j_n}$, the first event after the branch structure is $e_{end}$. Then we have,

$$
\begin{array}{l}
O'(e_b)<O'(e_{i_1})<O'(e_{i_2})<\dots <O'(e_{i_m})<O'(e_{end})\\
\wedge\\
O'(e_b)<O'(e_{j_1})<O'(e_{j_2})<\dots <O'(e_{j_m})<O'(e_{end})
\end{array}
$$

The order constraints capture the program order, which are unconditionally valid. That is also why we do not specify the order constraints under specific branch conditions.

{\bf Inter-thread value constraints\ } The inter-thread value constraints are identical to those in Section~\ref{sec:relax1}. Interesting, we do not need to specify the value constraints under the specific branch condition. The underlying reason is that, the shared read  $R$ in the instruction $x=R$ is isolated (meaning the value cannot be propagated downstream) if the instruction is disabled in the intra-thread value constraints due to the infeasible branch condition.
